---
title: "Lab 2 - Basic Data Management and Manipulation"
output: html_document
---

> **Scenario:** You are a data assistant to a newsroom editor. They want a **one-page brief** answering:  
> 1) How has U.S. voter turnout changed since 1980?  
> 2) Which measure— **VEP turnout** or **VAP turnout**—better matches ANES self-reports?  
> 3) How do **presidential** and **midterm** elections compare on VEP turnout?  

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = "##", collapse = TRUE)
```

---

The R Markdown file for this lab can be found here.
The data for this lab can be found [here]().

## Part A — Loading Data

Load the data into a data frame `turnout` and quickly verify its structure. Keep the vector of `year` and the list of column names.

```{r}
# YOUR CODE HERE
# turnout <- read.csv("turnout.csv")
# years <- turnout$year
# cols <- names(turnout)
# n_rows <- nrow(turnout); n_cols <- ncol(turnout)
# c(n_rows, n_cols); cols
```

---

## Part B — Construct the two turnout measures

Compute **VAP-based turnout** and **VEP-based turnout** (percent). Label both with the corresponding year so your later subsetting is self-documenting.

- VAP turnout: total votes divided by (VAP + overseas) × 100  
- VEP turnout: total votes divided by VEP × 100

Name your vectors `VAPtr`, `VEPtr`.

```{r}
# YOUR CODE HERE
# VAPtr <- (turnout$total / (turnout$VAP + turnout$overseas)) * 100
# VEPtr <- (turnout$total / turnout$VEP) * 100
# names(VAPtr) <- years
# names(VEPtr) <- years
```

---

## Part C — Reliability vs. ANES

Create the gaps between ANES self-reported turnout and each measure: `gapVAP` and `gapVEP`. Summarize both and **decide which measure aligns better**.

```{r}
# YOUR CODE HERE
# gapVAP <- turnout$ANES - VAPtr
# gapVEP <- turnout$ANES - VEPtr
# sumVAP <- summary(gapVAP)
# sumVEP <- summary(gapVEP)
# better_measure <- "VEPtr" # or "VAPtr" — choose based on your summaries
```

---

## Part D — Presidential vs. Midterms using indices

Define index vectors for **presidential** (`pres`) and **midterm** (`mids`) rows (years alternate). Subset `VEPtr` into `pVEPtr` and `mVEPtr`. Compute average VEP turnout for each and the difference.

```{r}
# YOUR CODE HERE
# pres <- c(1,3,5,7,9,11,13,14)  # indices for presidential years in this data
# mids <- c(2,4,6,8,10,12)       # indices for midterms
# pVEPtr <- VEPtr[pres]; names(pVEPtr) <- years[pres]
# mVEPtr <- VEPtr[mids]; names(mVEPtr) <- years[mids]
# vepp_diff <- mean(pVEPtr) - mean(mVEPtr)
```