---
title: "Lab 2 - Basic Data Management and Manipulation"
output: html_document
---

> **Scenario:** You are a data assistant to a newsroom editor. They want a **one-page brief** answering:  
> 1) How has U.S. voter turnout changed since 1980?  
> 2) Which measure— **VEP turnout** or **VAP turnout**—better matches ANES self-reports?  
> 3) How do **presidential** and **midterm** elections compare on VEP turnout?  

```{r setup, include=FALSE}
knitr::opts_chunk$set(include=FALSE)
```

---

The R Markdown file for this lab can be found here (link posted after class).
The data for this lab can be found [here](https://github.com/yuanzhouwang/yzwang-teaching).

## Part A — Loading Data

Load the data into a data frame `turnout` and quickly verify its structure. Keep the vector of `year` and the list of column names.

```{r}

```

---

## Part B — Construct the two turnout measures

Compute **VAP-based turnout** and **VEP-based turnout** (percent). Label both with the corresponding year so your later subsetting is self-documenting.

- VAP turnout: total votes divided by (VAP + overseas) × 100  
- VEP turnout: total votes divided by VEP × 100

Name your vectors `VAPtr`, `VEPtr`.

```{r}

```

---

## Part C — Reliability vs. ANES

Create the gaps between ANES self-reported turnout and each measure: `gapVAP` and `gapVEP`. Summarize both and **decide which measure aligns better**.

```{r}

```

---

## Part D — Presidential vs. Midterms using indices

Define index vectors for **presidential** (`pres`) and **midterm** (`mids`) rows (years alternate). Subset `VEPtr` into `pVEPtr` and `mVEPtr`. Compute average VEP turnout for each and the difference.

```{r}

```