---
title: "Lab 5: Correlations and Visualizations"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Install if needed
if (!requireNamespace("palmerpenguins", quietly = TRUE)) {
  install.packages("palmerpenguins")
}
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  install.packages("ggplot2")
}
if (!requireNamespace("dplyr", quietly = TRUE)) {
  install.packages("dplyr")
}
library(palmerpenguins)
library(ggplot2)
library(dplyr)
```

You'll also want to reference [this](https://github.com/yuanzhouwang/yzwang-teaching/blob/main/CSSS321%20ggplotlab.Rmd) tutorial on ggplot.

# Dataset

We'll use the **palmerpenguins** dataset, which contains measurements for three penguin species (Adelie, Chinstrap, Gentoo) from the Palmer Archipelago, Antarctica. Key variables include:

- `species` (factor: Adelie, Chinstrap, Gentoo)  
- `island` (factor)  
- `bill_length_mm`, `bill_depth_mm`  
- `flipper_length_mm`  
- `body_mass_g`  
- `sex` (male/female, with some missing)  

We'll focus on **flipper length** (in millimeters) as the outcome of interest and compare across species.

```{r packages}
peng <- penguins
peng <- peng %>% filter(!is.na(flipper_length_mm)) # drop NAs in outcome when needed
```

# 1. What is the **sample average flipper length** among **Gentoo** penguins?

```{r}
# [Your Code Here]
tapply(peng$flipper_length_mm, peng$species, mean)
mean(peng$flipper_length_mm[peng$species == "Gentoo"])
```

# 2. What is the **sample average flipper length** among **Adelie** penguins?

```{r}
# [Your Code Here]
```

# 3. What is the estimated difference in average flipper length of **Gentoo vs Adelie**? Is this effect Causal?

```{r}
# [Your Code Here]
mean(peng$flipper_length_mm[peng$species == "Gentoo"]) - mean(peng$flipper_length_mm[peng$species == "Adelie"]) %>%
  abs()
```


# 4. Visualize the distribution of `flipper_length_mm`. About how many observations are near **200 mm** (±5 mm)?

## Tidyverse (ggplot2) histogram
```{r}
# [Your Code Here]
ggplot(data = peng, mapping = aes(x = flipper_length_mm)) +
  geom_histogram(binwidth = 10)
```

## Base R histogram
```{r}
# hist(peng$flipper_length_mm, breaks = seq(floor(min(peng$flipper_length_mm, na.rm = TRUE)),
#                                           ceiling(max(peng$flipper_length_mm, na.rm = TRUE)), by = 5),
#      main = "Distribution of Flipper Length (base R)",
#      xlab = "Flipper length (mm)")
```

## Count near 200 mm

```{r}
peng %>% filter(abs(flipper_length_mm -  200) < 5) %>%
  nrow()

```

# 5. Explore the relationship between `flipper_length_mm` and `body_mass_g`. Does it look positive or negative?

## Tidyverse (ggplot2) scatter with trend
```{r}
ggplot(peng, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() +
  geom_smooth(method = "lm")

ggplot(peng, aes(y = flipper_length_mm, x = body_mass_g)) +
  geom_point() +
  geom_smooth(method = "lm")
```

## Base R scatter with fitted line
```{r}
plot(peng$flipper_length_mm, peng$body_mass_g)
```

# 6. Compute the correlation between `flipper_length_mm` and `body_mass_g`.

```{r}
cor(peng$flipper_length_mm, peng$body_mass_g)
```

**(a)** Are you surprised by the sign?
**(b)** Are you surprised by the absolute value (magnitude)?

# 7. If we wanted our estimates to generalize beyond the observed sample of penguins to all penguins of these two species, how should we select our sample?
